---
title: "Version Control with Git and Collaboration with GitHub"
format: revealjs
---

## Check link between R and GitHub {#check-link-between-r-and-github}

Configure your `user.name` and `user.email` for git in RStudio with:

```{r}
#| eval: false
usethis::use_git_config(
  user.name = "Jane Doe", # actual first and last name
  user.email = "jane@example.com" # email associated with GitHub account
)
```

Communicating with GitHub requires authentication with your GitHub account.
This is achieved by creating and securely storing a Personal Access Token (PAT).
The [Managing Git(Hub) Credentials](https://usethis.r-lib.org/articles/articles/git-credentials.html) vignette in the **`usethis`** [@usethis] package has details on using PATs, but for now the following should suffice.

Create a PAT with:

```{r}
#| eval: false
usethis::create_github_token()
```

-   Give the token a descriptive name, accept the default scopes, and click "Generate Token".
-   Copy the newly generated PAT to your clipboard.
-   Run `gitcreds::gitcreds_set()` in R and paste in your PAT when prompted.
-   After you close the PAT webpage, you won't be able to see your PAT again! You can store it somewhere secure, such as in a password manager.

### Caveat about storing PATs on Linux:

macOS and Windows both have built-in credential managers that will securely store your PAT, however the same does not exist in Linux.
[Happy Git with R has a section on options for storing your PAT on Linux](https://happygitwithr.com/https-pat#pat-doesnt-persist-on-linux).

::: aside
You may need to install the **`gitcreds`** package [@gitcreds].
:::

You can now check your Git setup by asking for a "situation report":

```{r}
#| eval: false
usethis::git_sitrep()
```

::: cell-output
```
── Git global (user)
• Name: 'Andy Teucher'
• Email: 'andy.teucher@gmail.com'
• Global (user-level) gitignore file: '/Users/andy/.gitignore'
• Vaccinated: TRUE
ℹ Defaulting to 'https' Git protocol
• Default Git protocol: 'https'
• Default initial branch name: 'main'

── GitHub user
• Default GitHub host: 'https://github.com'
• Personal access token for 'https://github.com': '<discovered>'
• GitHub user: 'ateucher'
• Token scopes: 'gist, repo, user, workflow'
• Email(s): 'andy.teucher@gmail.com (primary)'
ℹ No active usethis project
```
:::

The output shows information about the link between RStudio and GitHub.
Not all of it is relevant now, but it is essential to check whether your name and email are correct and the PAT is showing as "discovered":

::: cell-output
```
Personal access token for 'https://github.com': '<discovered>'
```
:::
